Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-03 manuscripts/manuscript.Rmd
edit all equations: time as function argument, not subscript. fix p vs theta, fix X_o X_p, fix K_ij

We let $X_t$ denote the size (numbers or biomass) of the focal population at time $t$ and assume that in the absence of take its dynamics are.

\begin{equation}
X_{t+1}= Z_t f(X_t) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $p$ is a vector of parameters to be estimated from the data.
We describe three choices for $f(X_t,p)$ in the next section.In this
simple model, the control action is a harvest or take, $h_t$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S_t=X_t-h_t$.

To construct the value function, we consider a return when $X_t=x$ and
harvest $h_t=h$ denoted as the reward, $R(x,h)$. For example, if the
return is simply the harvest at time $t$, then $R(x,h)=min(x,h)$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $\mathbf{E}_{X_t+1} [ \sum_{t = 0}^{T}  R_t(X_t, h_t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the
expectation over the process stochasticity in the future population state.

In order to find that policy, we introduce the value function $V_t(x_t)$
representing the total discounted catch from time $t$ onwards
given that $X_t=x_t$.  This value function satisfies an equation of
SDP  [@Mangel1988; @Clark2000; @Clark2009], sometimes called the
Hamilton-Jacobi-Bellman equation [@Mangel2014],

\begin{equation}
V_t(x_t) = \max_{h_t} \lbrace R(h_t, x_t) + \delta \cdot \mathbf{\mathrm{E}}_{X_{t+1}} \left[ V_{t+1}( X_{t+1}) | x_t, h_t \right] \rbrace
\end{equation}

where expectation is taken over all possible values of the next state,
$X_{t+1}$, and maximized over all possible choices of harvest, $h_t$.
That is, at time $t$, when population size is $x_t$ and harvest $h_t$ is
applied, the immediate return is $R(h_t, x_t)$. In the classical case,
the sole source of uncertainty is the process stochasticity term, $Z$,
and thus the expectation above is equivalent to taking expectations over
$Z$. That is

$$ \mathbf{\mathrm{E}}_{X_{t+1}} \left[ V_{t+1}( X_{t+1}) | x_t, h_t \right] = \mathbf{\mathrm{E}}_{Z} \left[ V_{t+1}( Z f(x_t - h_t))  | x_t, h_t \right] $$

where the population size after the take is $x_t-h_t$, which is then
translated into $X_{t+1}$ by Eqn 1 (that is, we just replaced $X_{t+1}$ with $Z f(x_t-h_t)$).

@@

$$\mathbf{\mathrm{E}}_{X_t+1} \left[ V_{t+1}( X_{t+1}) | x_t, h_t \right] = \mathbf{\mathrm{E}}_{\theta|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z | \theta, \mathrm{data}} \left[ V_{t+1}( Z f(x_t - h_t))  | x_t, h_t \right] \}$$

@@

$$\mathbf{\mathrm{E}}_{X_t+1} \left[ V_{t+1}( x_{t+1}) | X_t, h_t \right] = \mathbf{\mathrm{E}}_{\theta|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z | \theta, \mathrm{data}} \left[ V_{t+1}( Z f(x_t - h_t))  | x_t, h_t \right] \}$$

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [@Allen2005], the Myers
model [@Myers1995], Eqns \eqref{ricker}-\eqref{myers}, which we parameterize as follows for $f(S_t,p)$ in Eqn 1. In all three, we let $K$ denote the carrying capacity and $r$ the maximum per capita growth rate. In the Ricker model,

\begin{equation}
f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K} \right) } \label{ricker}
\end{equation}

@@

\begin{equation}
f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - X_c\right)} \label{allen}
\end{equation}

@@

\begin{equation}
f(S(t) | p)  = \frac{r S_t^{\theta}}{1 + \frac{S_t^\theta}{K}} \label{myers}
\end{equation}

@@

We assume that the data $X_o$ are observed with some process stochasticity,

$$X_{t+1} = f(X_t) + \varepsilon,$$

@@

Then under the GP, given a vector of observed data $X_o$ the predicted data $X_p$ obeys

$$f(X_p|X_o, X_{t+1}) \sim \mathcal{N}(E,C)$$
$$E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)^{-1} X_{t+1}$$
$$C = K(X_p, X_p) - K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)^{-1} K(X_o, X_p)$$

Where $K$ is the Gaussian process kernel.  (Compare to  Eqn 2.22 of @Rasmussen2006 for a more detailed derivation.) Throughout, we use the radial basis function kernel:

$$ K(\mathbf{x},\mathbf{y}) = \exp\left(\frac{-|\mathbf{x}-\mathbf{y}|^2}{2 \ell^2} \right)$$

@@

$$f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{-\alpha - 1}\exp\left(-\frac{\beta}{x}\right)$$

@@

Using the simulated training data we also estimate a Gaussian process
defined by a radial basis function kernel of two parameters: $\ell$,
which gives the characteristic length-scale over which correlation between
two points in state-space decays, and $\sigma$, which gives the scale
of the process noise by which observations $Y_{t+1}$ may differ from
their predicted values $X_{t+1}$ given an observation of the previous
state, $X_t$. @Munch2005a give an accessible introduction to the use
of Gaussian processes in providing a Bayesian nonparametric description
of the stock-recruitment relationship.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. The only difference 
is that the uncertainty in the future state under the GP, $X_{t+1}$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and "structural uncertainty" of the posterior collection of curves.  
(Recall that given the parameters, a parametric model is represented
by a single curve, while the GP remains a distribution of curves).  
From the GP posteriors, we can write down the (discritized) transition
matrix representing the probability of going to each state $X_{t+1}$ 
given any current state $X_t$ and any harvest $h_t$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE}
pandoc.table(rbind(dictable, aictable, bictable), caption="Model selection scores for several common criteria all (wrongly) select the simplest model. As the true (Allen) model is not distinguishable from the simpler (Ricker) model in the region of the observed data, this error cannot be avoided regardless of the model choice criterion. This highlights the danger of model choice when the selected model will be used outside of the observed range of the data.")
```

@@

```{r figure_3, dependson = c("assemble-opt"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979].", fig.width=8, fig.height=6, cache=FALSE}
policies <- melt(data.frame(stock=x_grid, sapply(OPT, function(x) x_grid[x])), id="stock")
names(policies) <- c("stock", "method", "value")

ggplot(policies, aes(stock, stock - value, color=method)) +
  geom_line(lwd=1.2) + 
  facet_wrap(~method) +
  xlab("stock size") + 
  ylab("escapement")  +
  scale_colour_manual(values=colorkey, guide=FALSE)

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3).
Policies are shown in terms of target escapement, $S_t$.  Under models
such as this a constant escapement policy is expected to be optimal
[@Reed1979], whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return the
population to $S$. Whenever a model predicts that the population will not persist below a certain threshold, the optimal solution is to harvest the entire population immediately, resulting in an escapement $S=0$, as seen in the true (correct form, exact parameters) model, the Allen model (correct form, estimated parameters) and the GP.  
Only the structurally correct model (Allen model) and the GP produce
policies close to the true optimum policy.

```{r figure_4, dependson=c("sim"), fig.cap="In the management context, GPDP outperforms approaches based on parametric models. We show 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.", fig.width=6, fig.height=6, cache=FALSE}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=0.1) +
  scale_colour_manual(values=colorkey, guide = FALSE) + 
  facet_wrap(~method) + 
  guides(legend.position="none", color=FALSE)

@@

$$P(x_t < x_c | t \in T) \propto  \exp\left(\frac{L^2}{\sigma^2}\right),$$

>>>>

We let $x(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are. (Throughout, we will use $X(t+1)$ with a capital $X$ as
a reminder that this is a stochastic, uncertainty quantity at time $t$,
while we use lowercase $x(t)$ as a reminder that the current population
size is treated as a given without error.  The latter assumption is
straight-forward to relax but will simplify the presentation here).

\begin{equation}
X(t+1) = Z(t) f(x(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $p$ is a vector of parameters to be estimated from the data.
We describe three choices for $f(X(t),p)$ in the next section.In this
simple model, the control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x $ and
harvest $h(t) = h$ denoted as the reward, $R(x, h)$. For example, if the
return is simply the harvest at time $t$, then $R(x, h) = \min(x, h)$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $\mathbf{E}_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the
expectation over the process stochasticity in the future population state.

In order to find that policy, we introduce the value function $V(x(t), t)$
representing the total discounted catch from time $t$ onwards
given that $X(t) = x(t)$.  This value function satisfies an equation of
SDP  [@Mangel1988; @Clark2000; @Clark2009], sometimes called the
Hamilton-Jacobi-Bellman equation [@Mangel2014],

\begin{equation}
V(x(t), t) = \max_{h(t)} \lbrace R(h(t), x(t)) + \delta \cdot \mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] \rbrace
\end{equation}

where expectation is taken over all possible values of the next state,
$X(t+1)$, and maximized over all possible choices of harvest, $h(t)$.
That is, at time $t$, when population size is $x(t)$ and harvest $h(t)$ is
applied, the immediate return is $R(h(t), x(t))$. In the classical case,
the sole source of uncertainty is the process stochasticity term, $Z$,
and thus the expectation above is equivalent to taking expectations over
$Z(t)$. That is

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V( X(t+1),t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{Z(t)} \left[V( Z(t) f(x(t) - h(t))), t+1 | x(t), h(t) \right]
\end{equation}

where the population size after the take is $x(t) - h(t)$, which is then
translated into $X(t+1)$ by Eqn 1 (that is, we just replaced $X(t+1)$ with $Z(t) f(x(t)-h(t))$).

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)), t)  \right] \}$$
\end{equation}

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z(t) | p, \mathrm{data}} \left[ V( Z(t) f(x(t) - h(t)), t) \right] \}
\end{equation}

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [@Allen2005], the Myers
model [@Myers1995], Eqns \eqref{ricker}-\eqref{myers}, which we parameterize as follows for $f(S(t),p)$ in Eqn 1. In all three, we let $K$ denote the carrying capacity and $r$ the maximum per capita growth rate. In the Ricker model,

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K} \right) } \label{ricker}
\end{equation}

@@

\begin{equation}
f(S(t)|p) = S(t) e^{r \left(1 - \frac{S(t)}{K}\right)\left(S(t) - X_c\right)} \label{allen}
\end{equation}

@@

\begin{equation}
f(S(t) | p)  = \frac{r S(t)^{\theta}}{1 + \frac{S(t)^\theta}{K}} \label{myers}
\end{equation}

@@

We assume that the data $X(t)$ are observed with some process stochasticity,

\begin{equation}
X(t+1) = g(X(t)) + \varepsilon,
\end{equation}

@@

We consider the fase where we have some previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm obs}(t), X_{\textrm obs}(t+1)$ for $t \in [1,2, ..., T]$

Under the GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau in [T+1, T+2, ..., T+n]$, is simply given by a multivariate normal ($\mathcal{N}$):

\begin{equation}
g(X_p|X_{\textrm obs}(t), X_{\textrm obs}(t+1)) \sim \mathcal{N}(E,C)
\end{equation}

which has mean and covariance matrix given by:

\begin{equation}
E = K(X_{\textrm p}(\tau), X_{\textrm obs}(t)) \left(K(X_{\textrm obs}(t),X_{\textrm obs}(t)) - \sigma \mathbb{I} \right)^{-1} X_{\textrm obs}(t+1)
\end{equation}

\begin{equation}
C = K(X_{\textrm p}(\tau), X_{\textrm p}(\tau)) - K(X_{\textrm p}(\tau), X_{\textrm obs}(t)) \left(K(X_{\textrm obs}(t),X_{\textrm obs}(t)) - \sigma \mathbb{I} \right)^{-1} K(X_{\textrm obs}(t), X_{\textrm p}(\tau))
\end{equation}

Where $K$ is the chosen covariance kernel.  (Compare to  Eqn 2.22 of @Rasmussen2006 for a more detailed derivation.) Many options are possible for this kernel. Throughout, we use the radial basis function for simplicity, a matrix whose $(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left(\frac{-|\mathbf{x_i}-\mathbf{j}|^2}{2 \ell^2} \right)
\end{equation}

@@

\begin{equation}
f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{-\alpha - 1}\exp\left(-\frac{\beta}{x}\right)
\end{equation}

@@

Using the simulated training data we also estimate a Gaussian process
defined by a radial basis function kernel of two parameters: $\ell$,
which gives the characteristic length-scale over which correlation between
two points in state-space decays, and $\sigma$, which gives the scale
of the process noise by which observations $Y(t+1)$ may differ from
their predicted values $X(t+1)$ given an observation of the previous
state, $x(t)$ [see @Munch2005a].

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. The only difference 
is that the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and "structural uncertainty" of the posterior collection of curves.  
(Recall that given the parameters, a parametric model is represented
by a single curve, while the GP remains a distribution of curves).  
From the GP posteriors, we can write down the (discritized) transition
matrix representing the probability of going to each state $X(t+1)$ 
given any current state $x(t)$ and any harvest $h(t)$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE}
pandoc.table(rbind(dictable, aictable, bictable), caption="Model selection scores for several common criteria (DIC: Deviance Information Criterion, AIC: Akaike Information Criterion, BIC: Bayesian Information Criterion) all (wrongly) select the simplest model. As the true (Allen) model is not distinguishable from the simpler (Ricker) model in the region of the observed data, this error cannot be avoided regardless of the model choice criterion. This highlights the danger of model choice when the selected model will be used outside of the observed range of the data.")
```

@@

```{r figure_3, dependson = c("assemble-opt"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S(t)$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979].", fig.width=8, fig.height=6, cache=FALSE}
policies <- melt(data.frame(stock=x_grid, sapply(OPT, function(x) x_grid[x])), id="stock")
names(policies) <- c("stock", "method", "value")

ggplot(policies, aes(stock, stock - value, color=method)) +
  geom_line(lwd=1.2) + 
  facet_wrap(~method) +
  xlab("stock size, x(t)") + 
  ylab("escapement, S(t)")  +
  scale_colour_manual(values=colorkey, guide=FALSE)

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3).
Policies are shown in terms of target escapement, $S(t) = x_t - h$.  Under models
such as this a constant escapement policy is expected to be optimal
[@Reed1979], whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return the
population to $S$. Whenever a model predicts that the population will not persist below a certain threshold, the optimal solution is to harvest the entire population immediately, resulting in an escapement $S=0$, as seen in the true (correct form, exact parameters) model, the Allen model (correct form, estimated parameters) and the GP.  
Only the structurally correct model (Allen model) and the GP produce
policies close to the true optimum policy.

```{r figure_4, dependson=c("sim"), fig.cap="In the management context, GPDP outperforms approaches based on parametric models. We show 100 replicate simulations of the stock dynamics (Eqn 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.", fig.width=6, fig.height=6, cache=FALSE}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=0.1) +
  scale_colour_manual(values=colorkey, guide = FALSE) +
  xlab("stock size, x(t)") +
  facet_wrap(~method) + 
  guides(legend.position="none", color=FALSE)

@@

$$P(x(t) < x_c | t \in T) \propto  \exp\left(\frac{L^2}{\sigma^2}\right),$$
